Interpretable Machine Learning

### 2020-09-03 2020-ecml-xkdd/paper/paper.tex
first bad draft

\paragraph{A Very Short History of IML.}
Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
Statistics has a long tradition of learning interpretable prediction  models from data, starting in 1800 \cite{stigler1986history} with works on least squares by Legendre 1805, \cite{legendre1805nouvelles}, Gauss in 1809 \cite{gauss1809theoria} and Quetelet in 1827 \cite{quetelet1827recherches}.
Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
  This took off in the 80s with decision tree approaches such as CART CITE, ID3 CITE and rule learners like AQ CITE but reaches back to first research in the 1960s \cite{hajek1966guha}.
Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g. model-based trees CITE).
As a next stepping stone to the field of IML are random forests, an early approach to explain more complex machine learning models.
I argue that one of the success factors of the random forest as a prediction model was (and is) the built-in feature importance measure.
The high citation counts of the original paper (CITE), but also suggestions for how to improve the feature importance (CITE, CITE, CITE) are evidence for this.
The IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (see Figure XXX) and papers published with these terms (see Figure XXX).
Could it be that only the terms have changed, but the field has always been there and the same?
I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time, see chapter (XXX).

@@

In this section I want to very broadly talk about methods that we have to explain models.
A first useful distinction is between inherently interpretable models and usually model-agnostic explanation methods that describe the behavior of the model.
This is a distinction between trying to make sense of a model by analyzing its parts vs. making sense of it by analyzing its prediction for certain inputs (sensitivity analysis).
However we interpret a model, the interpretation is always projected back to the features themselves or some abstract transformation of them.

\paragraph{Inherently interpretable models} are models where we can interpret its learned structure and parameters.
In this category, I see all linear regression models of the form $Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$, where the weights $\beta$ are learned parameters of the model.
Through the models structure (weighted sum of features) we can interpret the weights as the effect that a feature has on the prediction.
Similarly, decision trees and other rule-based ML models have a learned structure which we can interpret as how the model makes a prediction.
For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
However, the more complex these structures get (e.g., a linear model with hundres of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
Surprisingly, we can also try assing meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
This leads to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.

\paragraph{Sensitivity-analysis-like (model-agnostic) interpretation methods} often treats the model as a black-box (but not always).
Here I also want to distinguis between local and global explanations.

@@

\paragraph{Surrogate Models} are an interesting and huge class of explanation method.
This approach is very common in sensitivity analysis where we have emulation? models.
Surrogate models combine the two strategies:
First, approximate the predictions of the ML model with an interpretable model, such as a decision tree.
This is a kind of sensitivity-like approach, where we see the original model as a black box and artificially probe the ML model and derive information from the behavior.
Then we interpret the interpretable model, which now acts as a surrogate to the more complex ML model.

@@

\section{old}

\paragraph{A Very Short History of Machine Learning Interpretability}
Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
In statistics, there is a long tradition of learning interpretable models from data, starting in 1800 \cite{stigler1986history}.
The least squares  method to fit linear models reaches back to 1805 by Legendre. \cite{legendre1805nouvelles}
Later the linear regression model became a tool for astronomy and geodesy, e.g. by Gauss in 1809 \cite{gauss1809theoria}.
Starting 1827 Quetelet used linear regression models (with least squares method) for (at that time) large amounts of data from social sciences, such as crime data \cite{quetelet1827recherches}.
We can hardly call the least squares method interpretable machine learning, but it remains a highly relevant building block.
Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
Until 1950s mostly statistical methods available, so I guess we can't speak yet of interpretable machine learning.
A side note: I don't want to distinguish too sharply between statistics and machine learning.
But for this context, I would say in statistical modeling the focus is on finding a suitable linear model using in-sample metrics such as R squared and AIC.
In machine learning we use an external test set to minimize the generalization loss and the model itself does not matter as long as the loss is low.

I want to highlight the random forest CITE as a representative for early interpretable ML approach, which also lay out some more groundwork and a new paradigm for feature importance.

In 2010s then came the deep learning hype.

I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.

The big leap came in the mid 2010's, visible in Google Search trends [link] and Web of Science statistics when searching for 'Explainable AI' and 'Interpretable Machine Learning'.
These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion.

Research was done, big time in statistics for regression models in its various extensions, rule based stuff in computer science, random forest and its many extensions and interpretation.
What has changed is that Interpretable Machine Learning or Explainable Artificial Intelligence has become a field that now ties all these topics together.
Especially model-agnostic methods that work for any type of model have grown a lot in this new field.

\paragraph{Where Do We Stand Today in 2020?}

We have software that implements interpretability (CITE lots of stuff here).
Many of the big IT companies
We have software, big companies that offer solutions and many startups.
Research is heavily underway and people are already using the solutions.

\paragraph{What Methods do We Have?}

There are many overview papers with categorizations of all these methods.
Here are the properties I find the most useful to categorize the various methods and allow a user to understand when to use which.

An interesting distinction is how the methods work:
They either explain individual 'parts' of the model, examples are the weights in a linear model, the structure of a tree.
This can be very straightforward such as the interpretation of a linear regression weight, which directly tells us how a feature influences the response.
Or this can be more complex, when we want to understand what the role of individual neurons in a convolutional neural network is.
Here we can do activation maximization which tells us to which image the neuron most strongly reacts to and interpret this neuron as, e.g., a detector of a certain pattern.
Methods that require to look at certain parts of the model are by definition model-specific, i.e. they cannot be applied to any model.
Methods that mostly observe the input - output behavior are often model-agnostic, i.e. they ignore whether the model is a neural network, linear model or random forest.

A second distinction is about whether a method explains an individual data point or the global model behavior.
Methods such as Shapley Values and LIME compute explanations for points, while there are many feature importance methods that quantify the importance of a feature globally for the model, meaning across all data points.

\subsection{Linear Models and Rule-based ML}

Linear models and rule can be seen as building blocks for many types of explanations and explanation methods.
Interpretations are often presented as a weighted sum of the inputs, such as LIME, Shapley Values, and many of the attribution methods for convolutional neural networks.
Even the entire model can -- in theory -- be deconstructed into and additive sum of components reaching from first-order effects to all interactions (cite functional ANOVA).
But also rule based explanations are very common types of explanations.
There are many new tree algorithms, such as XXX and XXX.

\paragraph{Linear Models}
The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
That makes estimation and interpretation simple.
But this is often not adequate since in the real world the effects are usually not linear.
Many extensions exists, such as non-linear transformations for the case that the target follows a different distribtuion (GLMs, logistic regression).
Other extensions allow for non-linear effect of a feature, by using splines.
Also interactions can be integrated.
All these extensions make the model more difficult to interpret, but also allow to more flexible model the target.

\paragraph{Rule-based ML}
There are different approaches.
They all have in common that they find decision rules in the form of conditions combined with AND conjunctions leading to some prediction if all conditions are matched.
There are different ways to find this rules and if they are allowed to overlap and so on.
Differences in how they are optimized and so on.

\subsection{Explain Individual Predictions}

There was a great explosion in different methods to explain individual predictions of ML models in a model-agnostic way.
Among the most popular are Local Interpretable Model-agnostic Explanations (LIME) CITE, Shapley Values from Game Theory (CITE x2) and Counterfactual Explanations.
Their outputs of LIME and Shapley Value are weighted sums and are estimated using method from linear regression, again showing the deep roots with statistics.
Counterfactual explanations are presented as decision rules, for LIME this is also possible

\paragraph{Surrogate Models}
Many methods -- a confusing amount, really -- are nothing but simple models that approximate the behavior of the (complex) model to be explained.

\paragraph{Counterfactual Explanations}
Counterfactual explanations are rather simple explanations for individual predictions.
A counterfactual explanation tells you how the features should be changed to get to a different prediction or classification.
But when you look closer at how you find them, it gets more difficult.
There are many tradeoffs to make.
First, you want to get as close as possible to the desired outcome.
At the same time you don't want to change too many features, and the features that are changed should not be changed too much.
Also these changes should not create unrealistic combinations of feature values, think of a 2m person for which the weight was changed to 30kg.
The reason that there are many papers on the topic is that the search for counterfatuals can be expensive.
Some methods are only for specific types of models, and they all trade-off the various objectives in different ways.
We argued that it is necessary to treat this as a multi-objective problem and the result should be not one but many counterfactual explanations.

\paragraph{LIME}

\paragraph{Shapley Values}

\subsection{Global Model Interpretation}

\paragraph{Feature Effects}

\paragraph{Feature Importance}

\subsection{Other Methods}

\paragraph{Heatmaps}

@@

\subsection{Dependent Features}
Feature dependence comes in many forms, correlation being the simplest and most known.
When features are dependent -- and they usually are -- disentangling the effect of one from the other are difficult if not impossible.
For example, if you measure various markers of inflammation in a blood sample, they will be highly correlated.
Using these markers in a predictive model for classifying diseases for which these markers are highly predictive, the model can rely on any of them with arbitrary weighting.
Attributing which marker was the most important now becomes difficult.

Almost all interpretation methods are affected by this issue.
Most of them ignore this issue and pretend the features are independent of each other.
For example the permutation feature importance measure quantifies the importance as the drop in performance when the feature is permuted.
But permutation when features are dependent not only breaks the association with the target value, but also with the other features.
Now if we permute one blood marker, it might create new combination that might be very unlikely.
These combinations are now fed to the model to get the predictions, but these are outside of the training data distribution and the model never had to deal with similar combinations.
It is totally unclear how the model will behave and it might even be irrelevant how the model will behave for these data points as they will never occur in real life.
There are attempts to 'fix' this issue, namely to not permute the feature, but to draw from the conditional distribution.
This means we still sample new values, but only the ones that are conforming with the other feature values.
Unfortunately, this is not a simple 'fix' of the issue.
While it solves the extrapolation issue, it becomes a new measure altogether, since conditional feature importance measure the drop in performance "given that we know the other feature values".
This can mean that the two most important features can drop to almost zero importance when yhey are highly correlated.

Similar things happen to 'fixes' for e.g. Shapley Values.
Once we start sampling from the conditional distribution and not from the marginal, we don't get Shapley Values for our original question, but something else.

I fear that there is no easy solution.

\subsection{Causality}
In general, we are not allowed to make a causal interpretation of machine learning models.
There is an inherent conflict between prediction performance optimization and causal modeling.
A feature might give some additional predictive performance, but its effect may not be interpreted causally, because, e.g., we did not include the confounder of both this feature and the target.
The fact that the ground has some predictive value for tomorrow's weather, but it's not causal for it.
The common confounder is today's weather -- todays rain makes the ground wet.
And if we had no access to todays weather but only to the "wet ground"-feature, we have to decide between predictive performance and causal correct interpretation.

\subsection{Inference}
Example for confidence intervals for LRP https://arxiv.org/pdf/2008.01468.pdf.
Is model-specific for Deep Neural Networks using Dropout.

\subsection{Many More Topics}
This was not an exhaustive list, and especially focused on the methodological side, neglecting views of the user interface, societal impacts, psychological insights and so on.
These challenges are more from the methodological, mathematical view, i.e., what can be done given we have some data and model.
This is also a rather static view point in the sense that we can do a lot if we allow a more interactive way of modeling and "having a conversation" between a person and the model or even the modeling process.

@@

While Interpretable Machine Learning can seem like a pretty young field, it has some old roots in Statistics and Computer Science.
Yet, many new methods have come in the recent years and we have witnessed a Cambrian explosion of methods, software and institutional interest.
We are seing both ways of making models themselves more interpretable, but also model-agnostic methods that work for any type of model.
These methods also are implemented now in many software packages.
Big leaps in the field came from old ideas in other fields.
The Shapley Value has been for a long time part of game theory, and its introduction to machine learning interpretability has spurred a lot of development.
We can learn a lot from Statistics and how it handles uncertainty.

>>>>

TODO: Mention sensitivity analysis

\paragraph{A Very Short History of IML.}
Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
Statistics has a long tradition of learning interpretable prediction  models from data, starting in 1800 \cite{stigler1986history} with works on least squares by Legendre 1805, \cite{legendre1805nouvelles}, Gauss in 1809 \cite{gauss1809theoria} and Quetelet in 1827 \cite{quetelet1827recherches}.
Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
Many extensions exists, such as non-linear transformations for the case that the target follows a different distribtuion (GLMs, logistic regression).
The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
  This took off in the 80s with decision tree approaches such as CART CITE, ID3 CITE and rule learners like AQ CITE but reaches back to first research in the 1960s \cite{hajek1966guha}.
Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g. model-based trees CITE).
As a next stepping stone to the field of IML are random forests, an early approach to explain more complex machine learning models.
I argue that one of the success factors of the random forest as a prediction model was (and is) the built-in feature importance measure.
The high citation counts of the original paper (CITE), but also suggestions for how to improve the feature importance (CITE, CITE, CITE) are evidence for this.

In 2010s then came the deep learning hype.
And a few years after that, the IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (see Figure XXX) and papers published with these terms (see Figure XXX).
Could it be that only the terms have changed, but the field has always been there and the same?
I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time, see chapter (XXX).
These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion, since developments happened under different names and in different fields.
What has changed is that Interpretable Machine Learning or Explainable Artificial Intelligence has become a field that now ties all these topics together.

@@

In this section I want to very broadly talk about methods that we have to explain models.
However we interpret a model, the interpretation is always projected back to the features themselves or some abstract transformation of them.
Often the distinction is made between model-specific (works only for one type of model) and model-agnostic (works for all models) interpretation methods.
I want to distinguish between introspection and sensitivity analysis.
An interpretation method that works by introspection assigns semantic meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model and describes its behavior}.
Introspection is necessarily model-specific.
Sensitivity analysis is mostly model-agnostic, but can also be done by leveraging model-specific knowledge such as gradient-based methods. FOOTNOTE: An example are the various heatmap-like explanation for CNNs for image classification (CITE many) which study the sensitivity of the prediction to the input pixels. Technically this is done by backpropagation and therefore model-specific. Model-agnostic versions (CITE) exist.
The other useful category is local vs. global explanations.
Local means that the method explains individual predictions, and global means it explains the average model behavior.
For inherently interpretable models, global and local explanations are often the same: In linear models, the coefficients both how the model behaves globally, but also is used for individuals instances.

\paragraph{Inherently interpretable models} are models where we can interpret its learned structure and parameters.
As a rule of thumb for when we do an introspection is that we are not probing the model with various (manipulated data points).
In this category, I see all linear regression models of the form $Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$, where the weights $\beta$ are learned parameters of the model.
Through the models structure (weighted sum of features) we can interpret the weights as the effect that a feature has on the prediction.
Similarly, decision trees and other rule-based ML models have a learned structure which we can interpret as how the model makes a prediction.
For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
However, the more complex these structures get (e.g., a linear model with hundres of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
Surprisingly, we can also try assign meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
This leads to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
Another example is the random forest minimal depth distribution TODO: CITE
The distribution of minimal depth shows for a feature for all the trees in the forest at which depth it occurs, with the notation that it the feature is more important if it appears early in the tree, it is more important.
This is again an example how we analyze the structure of a model to get an understanding of how it makes decisions, but the model is more complex (hundreds of decision trees) so there is more work to interpreting its learned structure.
The random forest is also a useful example for another measure, the feature importance, for which various measures exist, such as decrease in accuracy and decrease in Gini impurity.
By using our distinction of introspection and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
Gini importance collects for each feature and tree how the split decreased the Gini impurity.
It does not rely on making predictions, but only on internal statistics.
For mean decrease accuracy, we probe the random forest with perturbed data, i.e., we shuffle a feature and see how the prediction changes.
It's a model-specific methods since we do this tree-wise and use the out-of-bag sample of the tree, so this procedure is cleverly adapted to the random forest.
Model-agnostic versions of this permutation feature importance exists ( TODO: CITE).

\paragraph{Sensitivity-analysis-like (model-agnostic) interpretation methods} often treats the model as a black-box (but not always).
Here I also want to distinguish between local and global explanations.

@@

\paragraph{Surrogate Models} are an interesting and huge class of explanation method.
This approach is very common in sensitivity analysis where we have emulation? models.
Many methods -- a confusing amount, really -- are nothing but simple models that approximate the behavior of the (complex) model to be explained.
Surrogate models combine the two strategies:
First, approximate the predictions of the ML model with an interpretable model, such as a decision tree.
This is a kind of sensitivity-like approach, where we see the original model as a black box and artificially probe the ML model and derive information from the behavior.
Then we interpret the interpretable model, which now acts as a surrogate to the more complex ML model.

@@

@@

There are also challenges for the field, that are already being tackled but still remain issues.
This list is also based on this paper (CITE Pitfalls).

\paragraph{Feature dependence} makes interpretation more difficult.
Inherently, when two features are dependent, e.g., correlated, attributing effects and importance to them is difficult.
Most methods also suffer from extrapolation when features are dependent.
When features are, e.g., permuted individually, this breaks the association with the target to predict (that is desirable) but it also breaks the association with the other features (which can cause problems).
The problem here is that new data points are being generated that are unlikely.
These points are then used for computing importance values, explanations for individual predictions and so on.
Many model-agnostic methods have this issue: PFI, PDP, Shapley values, LIME, ...
But also interpretation for inherently interpretable models can suffer:
When features are strongly correlated, the variance of estimated weights in a linear model can become large.
Also the interpretation of a weight is conditional that all other features remain fixed, which becomes the more unrealistic the more dependence there is between features.
There are some "fixes" for it, which enforce that any manipulation of the data respects the joint distribution.
All solutions have the problem that they now mix the effects and importance of features that switch the focus of interpretation or even break the original interpretation method in the case of Shapley values (CITE, CITE).
Solutions exist for feature importance, called conditional feature importance CITE APPROACHES.
Currently there is no satisfying solution that fixes extrapolation, but keeps the original marginal interpretation.
I fear that there is no easy solution.

\paragraph{Inference}
In statistics it is common procedure to quantify the uncertainties that come with estimates of, e.g., coefficients in a linear model.
Quantification of uncertainty, be it in the form of error bars, confidence intervals, test statistics or p-values, helps to separate signal from noise.
Many techniques in IML currently only give the point estimates, but don't quantify the uncertainty.
For things like feature importance there are variations, but we need to have these things built into common software tools.
Also it is unclear which uncertainty should be quantified and for what purpose.
Should we treat the model as fixed?
Or should we quantify the uncertainty of some feature effect in the same manner as we do for the generalization error, that is with for example cross-validation.
Here we have to dig deeper into the statistics literature which can be our guide.
Example for confidence intervals for LRP https://arxiv.org/pdf/2008.01468.pdf.
Is model-specific for Deep Neural Networks using Dropout.

\paragraph{Causal interpretation}
Usually, we desire a causal interpretation of an effect or importance.
As a researcher you want to draw conclusion about the true underlying process that generated your observations and learn about the world.
When you want to use ML for decision making, you need causal reasons that can change the target in the real world.
For example, when modeling the likelihood that customers will quit their subscription, only a good prediction will not be enough, but you have the desire to prevent a customer from leaving.
Only causal interpretation of the model will help you find reasonable real-world interventions.
However, this is generally not possible and causality can even be a conflicting goal to predictive accuracy.
For example when we don't have access to an important confounder of a (non-causal) feature and the target, it improve predictive performance if we include the non-causal feature.
When we want to predict whether it will rain tomorrow, a useful feature might be whether the ground is wet today, assuming we don't have access to today's weather.
Todays weather is the cause for both the wet ground and tomorrows weather.
Including it in the model will improve performance, but we are not allowed to interpret the effect and importance for the wet ground causally for the weather tomorrow.
Here, knowledge from the causality community is important.
It's not enough to somehow fix causality within the interpretation method, but we must go back to the modeling process itself to consider causality already there.

\paragraph{Lack of definition of interpretability and explainability} is a common critique of the field (TODO: CITE some papers, e.g. Lipton).
First, I don't think the lack of definition is a problem for the "naming" of the field.
I think of "Intepretable ML" and "Explainable AI" as useful terms to collect all topics that are related to analyzing ML models beyond performance metrics.
There is also very valid critique in lack of definition.
Because there is no single, clean definition of interpretability, quantitative research results can be hard to come by, e.g., how can you show that some interpretation method delivers better explanations than another?
This is connected to a lack of groundtruth, we just don't know what the correct explanation is.
To counter this argument, interpretability depends on audience and context (CITE PLEASE StOP?)
It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity (CITE), interaction strength (CITE) and much more (CITE ...).
It will take some time to establish some best practice for evaluation interpretation methods and the explanations they produce.

\paragraph{And much more} research has to be done on which I did not touch.
This was not an exhaustive list, and especially focused on the methodological side, neglecting views of the user interface, societal impacts, psychological insights and so on.
These challenges are more from the methodological, mathematical view, i.e., what can be done given we have some data and model.
This is also a rather static view point in the sense that we can do a lot if we allow a more interactive way of modeling and "having a conversation" between a person and the model or even the modeling process.
How do we present multiple maybe conflicting explanations, e.g., when you generate multiple counterfactual explanations?

@@

While Interpretable Machine Learning can seem like a pretty young field, it has some old roots in Statistics and Computer Science.
Yet, many new methods have come in the recent years and we have witnessed a Cambrian explosion of methods, software and institutional interest.
We are seeing both ways of making models themselves more interpretable, but also model-agnostic methods that work for any type of model.
These methods also are implemented now in many software packages.
Big leaps in the field came from old ideas in other fields.
The Shapley Value has been for a long time part of game theory, and its introduction to machine learning interpretability has spurred a lot of development.
We can learn a lot from Statistics and how it handles uncertainty.
